---
title: "Class–Survival Gap Analysis: Access vs Priority Rules"
author: "CLI Copilot"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
---

# Introduction

This report addresses the research question: **How much of the class–survival gap is explained by access and constraints rather than priority rules?**

## Analysis Plan

We will:
1. Explore the Titanic dataset and relevant variables (class, survival, access, constraints, priority rules).
2. Propose a stepwise analysis to decompose the class–survival gap.
3. For each code chunk, provide detailed explanation of the logic, assumptions, and relevance to the research question.

# Data Overview

```{r}
library(tidyverse)
library(knitr)
train <- read_csv("../../data/train.csv")
test <- read_csv("../../data/test.csv")
summary(train)
```

# Class–Survival Gap

```{r}
class_survival <- train %>%
  group_by(Pclass) %>%
  summarize(survival_rate = mean(Survived), n = n())
knitr::kable(class_survival)
```

# Decomposing the Gap: Access & Constraints vs Priority Rules

```{r}
# Placeholder for further decomposition analysis
# e.g., stratify by sex, age, embarkation, etc.
```

# Next Steps

- Refine decomposition logic
- Visualize findings
- Document all steps and reasoning
