---
title: "Genetic Algorithm for Feature Selection: Titanic Survival Prediction"
author: "CLI Copilot"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
---

# 1. Introduction and Approach

## Research Question
How can genetic algorithms be used to select optimal features from an expanded feature space (50 features) for predicting passenger survival on the Titanic using logistic regression?

## Methodology Overview
This analysis follows a systematic approach:

1. **Data Preparation**: Load and explore the Titanic training dataset
2. **Feature Engineering**: Create 50 candidate features through transformations and combinations of existing variables
3. **Feature Documentation**: Provide logical explanations for each proposed feature
4. **Genetic Algorithm Implementation**: Use a GA to identify the optimal subset of features
5. **Model Evaluation**: Train logistic regression models and compare performance before and after feature selection

## Key Assumptions
- The training data is representative of the underlying population
- Feature transformations maintain interpretability
- Logistic regression is a suitable baseline for survival prediction
- The genetic algorithm will find near-optimal solutions through evolutionary pressure on prediction accuracy

---

# 2. Data Loading and Exploration

```{r}
#| label: load-libraries
#| eval: true

# Load required libraries
library(tidyverse)
library(caret)
library(GA)
library(pROC)
library(knitr)

# Set random seed for reproducibility
set.seed(42)
```

We begin by loading all necessary libraries: `tidyverse` for data manipulation, `caret` for machine learning utilities, `GA` for genetic algorithms, `pROC` for ROC curve analysis, and `knitr` for table formatting. The random seed ensures reproducibility of stochastic processes in the genetic algorithm.

```{r}
#| label: load-data
#| eval: true

# Load the Titanic training dataset
titanic_raw <- read_csv('/Users/patrickl/Developer/Personal-Repo/cli_poc_ml_project/data/train.csv')

# Display basic information
dim(titanic_raw)
head(titanic_raw, 3)
colnames(titanic_raw)
```

The raw Titanic dataset contains `r nrow(titanic_raw)` observations and `r ncol(titanic_raw)` original variables. We load this data to understand its structure before feature engineering. The original features include demographic information (sex, age), socio-economic indicators (pclass, fare), family relationships (sibsp, parch), port information (embarked), and ticket/cabin details.

```{r}
#| label: examine-missingness
#| eval: true

# Check for missing values
missing_summary <- titanic_raw %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Missing_Count") %>%
  arrange(desc(Missing_Count))

kable(missing_summary, caption = "Missing Values by Variable")

# Get data types
data_types <- tibble(
  Variable = colnames(titanic_raw),
  Type = sapply(titanic_raw, class)
)
kable(data_types, caption = "Data Types of Original Variables")
```

The data contains missing values in Age (n=`r missing_summary$Missing_Count[missing_summary$Variable == 'Age']`), Cabin (n=`r missing_summary$Missing_Count[missing_summary$Variable == 'Cabin']`), and Embarked (n=`r missing_summary$Missing_Count[missing_summary$Variable == 'Embarked']`). Understanding these patterns is critical for feature engineering decisions.

---

# 3. Feature Engineering: Creating 50 Candidate Features

```{r}
#| label: feature-engineering
#| eval: true

# Create enhanced dataset with 50 engineered features
titanic_engineered <- titanic_raw %>%
  # Handle missing values first
  mutate(
    Age = replace_na(Age, median(Age, na.rm = TRUE)),
    Embarked = replace_na(Embarked, names(which.max(table(Embarked))))
  ) %>%
  # ORIGINAL FEATURES (keep some for baseline)
  mutate(
    # 1. Pclass: Ticket class (proxy for socio-economic status)
    f_pclass = as.numeric(Pclass),
    
    # 2. Sex: Gender (0=Female, 1=Male)
    f_sex_male = as.numeric(Sex == "male"),
    
    # 3. Age: Passenger age
    f_age = Age,
    
    # 4. Fare: Ticket price
    f_fare = Fare,
    
    # 5. SibSp: Siblings/spouses aboard
    f_sibsp = SibSp,
    
    # 6. Parch: Parents/children aboard
    f_parch = Parch,
    
    # 7. Embarked port (one-hot encoded)
    f_embarked_c = as.numeric(Embarked == "C"),
    f_embarked_q = as.numeric(Embarked == "Q"),
    f_embarked_s = as.numeric(Embarked == "S"),
    
    # ===== POLYNOMIAL & INTERACTION FEATURES =====
    # 8-10. Polynomial transformations of age
    f_age_squared = Age^2,
    f_age_cubed = Age^3,
    f_log_age = log(Age + 1),
    
    # 11-13. Fare transformations
    f_fare_squared = Fare^2,
    f_log_fare = log(Fare + 1),
    f_sqrt_fare = sqrt(Fare),
    
    # ===== FAMILY & SOCIAL FEATURES =====
    # 14. Total family size
    f_family_size = SibSp + Parch + 1,
    
    # 15. Has family aboard (binary indicator)
    f_has_family = as.numeric((SibSp + Parch) > 0),
    
    # 16. Is alone (no family)
    f_is_alone = as.numeric((SibSp + Parch) == 0),
    
    # 17. Has spouse aboard
    f_has_spouse = as.numeric(SibSp > 0),
    
    # 18. Has children aboard
    f_has_children = as.numeric(Parch > 0),
    
    # 19. Number of parents aboard
    f_num_parents = Parch,
    
    # 20. Ratio of family to class (socio-economic context)
    f_family_to_class = (SibSp + Parch) / Pclass,
    
    # ===== AGE-BASED FEATURES =====
    # 21. Is child (age < 18)
    f_is_child = as.numeric(Age < 18),
    
    # 22. Is young adult (18-35)
    f_is_young_adult = as.numeric(Age >= 18 & Age <= 35),
    
    # 23. Is middle aged (35-60)
    f_is_middle_aged = as.numeric(Age > 35 & Age <= 60),
    
    # 24. Is elderly (60+)
    f_is_elderly = as.numeric(Age > 60),
    
    # 25. Age deviation from class median
    f_age_class_deviation = Age - median(Age[Pclass == .$Pclass], na.rm = TRUE),
    
    # ===== FARE-BASED FEATURES =====
    # 26. Fare per family member
    f_fare_per_family = Fare / (SibSp + Parch + 1),
    
    # 27. Is high fare (top quartile)
    f_high_fare = as.numeric(Fare >= quantile(Fare, 0.75, na.rm = TRUE)),
    
    # 28. Is low fare (bottom quartile)
    f_low_fare = as.numeric(Fare < quantile(Fare, 0.25, na.rm = TRUE)),
    
    # 29. Fare relative to class median
    f_fare_class_ratio = Fare / (median(Fare[Pclass == .$Pclass], na.rm = TRUE) + 1),
    
    # ===== CLASS-BASED FEATURES =====
    # 30. Is first class
    f_is_first_class = as.numeric(Pclass == 1),
    
    # 31. Is second class
    f_is_second_class = as.numeric(Pclass == 2),
    
    # 32. Is third class
    f_is_third_class = as.numeric(Pclass == 3),
    
    # 33. Is upper class (1st or 2nd)
    f_is_upper_class = as.numeric(Pclass <= 2),
    
    # ===== INTERACTION FEATURES =====
    # 34. Female in first class (high priority rescue)
    f_female_first_class = f_sex_male == 0 & Pclass == 1,
    
    # 35. Female in second class
    f_female_second_class = f_sex_male == 0 & Pclass == 2,
    
    # 36. Male in first class
    f_male_first_class = f_sex_male == 1 & Pclass == 1,
    
    # 37. Child in first class
    f_child_first_class = f_is_child & Pclass == 1,
    
    # 38. Child and female
    f_child_female = f_is_child & f_sex_male == 0,
    
    # 39. Age × Class interaction
    f_age_class_interaction = Age * Pclass,
    
    # 40. Fare × Class interaction
    f_fare_class_interaction = Fare * Pclass,
    
    # ===== CABIN-BASED FEATURES =====
    # 41. Has cabin information
    f_has_cabin_info = as.numeric(!is.na(Cabin)),
    
    # 42. Extract cabin deck letter (A, B, C, D, E, F, G, T)
    f_cabin_deck = as.numeric(factor(substr(Cabin, 1, 1))),
    
    # ===== TICKET-BASED FEATURES =====
    # 43. Ticket starts with digit (different ticketing patterns)
    f_ticket_numeric_start = as.numeric(grepl("^[0-9]", Ticket)),
    
    # 44. Ticket length (categorical vs numeric patterns)
    f_ticket_length = nchar(Ticket),
    
    # ===== DERIVED RISK SCORES =====
    # 45. Female advantage score (females more likely to survive)
    f_female_score = (1 - f_sex_male) * (1/Pclass),
    
    # 46. Low class score (3rd class had lower survival)
    f_third_class_risk = f_is_third_class * 2,
    
    # 47. Age-gender combined score
    f_young_female_score = f_is_child * (1 - f_sex_male) * (1 - f_is_third_class),
    
    # ===== STATISTICAL FEATURES =====
    # 48. Z-score of age (standardized age within dataset)
    f_age_zscore = (Age - mean(Age, na.rm = TRUE)) / sd(Age, na.rm = TRUE),
    
    # 49. Z-score of fare (standardized fare within dataset)
    f_fare_zscore = (Fare - mean(Fare, na.rm = TRUE)) / sd(Fare, na.rm = TRUE),
    
    # 50. Survival label (target variable - keep for reference)
    target_survival = Survived
  ) %>%
  # Drop original columns, keep only engineered features
  select(starts_with("f_"), target_survival) %>%
  # Remove any constant or near-constant features
  select(where(~var(., na.rm = TRUE) > 0.001 | is.na(var(., na.rm = TRUE))))

# Verify final feature count
n_features <- ncol(titanic_engineered) - 1  # Exclude target
n_features
```

We have created `r n_features` engineered features from the original dataset. The feature engineering process combines mathematical transformations (polynomial, logarithmic, square root), domain-specific interactions (female in first class), and statistical transformations (z-scores).

---

# 4. Feature Documentation and Rationale

```{r}
#| label: feature-documentation
#| eval: true

# Create comprehensive feature documentation
feature_rationale <- tribble(
  ~Feature_Name, ~Category, ~Rationale_for_Prediction,
  
  # Original features
  "f_pclass", "Original", "Socio-economic status proxy; 1st class passengers had better access to lifeboats and information",
  "f_sex_male", "Original", "Gender was critical to survival; 'women and children first' evacuation protocol prioritized females",
  "f_age", "Original", "Age affects physical ability to navigate, reach lifeboats, and prioritization in evacuation",
  "f_fare", "Original", "Fare indicates socio-economic status, cabin location, and quality of accommodations affecting proximity to lifeboats",
  "f_sibsp", "Original", "Presence of siblings/spouses may affect survival decisions and evacuation coordination",
  "f_parch", "Original", "Presence of parents/children may force survival/death decisions (e.g., staying with family)",
  
  # Embarked features
  "f_embarked_c", "Embarked", "Port of embarkation (Cherbourg) may indicate different passenger demographics and wealth",
  "f_embarked_q", "Embarked", "Port of embarkation (Queenstown) associated with different passenger characteristics",
  "f_embarked_s", "Embarked", "Port of embarkation (Southampton) dominant embarkation point affects regional/class composition",
  
  # Polynomial & transformation features
  "f_age_squared", "Polynomial", "Non-linear relationship; very young and very old may have different survival patterns than middle-aged",
  "f_age_cubed", "Polynomial", "Captures extreme age effects with stronger power law relationship",
  "f_log_age", "Polynomial", "Log transformation reduces the impact of outlier ages while preserving patterns",
  "f_fare_squared", "Polynomial", "Fare has non-linear relationship with survival; high-fare passengers had premium accommodations",
  "f_log_fare", "Polynomial", "Log transformation emphasizes differences between lower-priced tickets",
  "f_sqrt_fare", "Polynomial", "Alternative scaling of fare capturing different aspects of wealth indicator",
  
  # Family features
  "f_family_size", "Family", "Larger families may have difficulty coordinating escape; affects evacuation speed and decision-making",
  "f_has_family", "Family", "Presence of family aboard affects survival strategy and evacuation compliance",
  "f_is_alone", "Family", "Solo travelers may have higher mobility and escape speed without dependent care",
  "f_has_spouse", "Family", "Spouse presence indicates married status; may affect evacuation behavior and priority",
  "f_has_children", "Family", "Children aboard create obligation affecting parent behavior in evacuation",
  "f_num_parents", "Family", "Presence of parents affects evacuation dynamics and dependency relationships",
  "f_family_to_class", "Family", "Interaction of family structure with socio-economic class reveals vulnerability",
  
  # Age-based features
  "f_is_child", "Age_Groups", "Children were prioritized in evacuation; universal survival advantage regardless of class",
  "f_is_young_adult", "Age_Groups", "Young adults may have better physical capabilities but no priority in evacuation",
  "f_is_middle_aged", "Age_Groups", "Middle-aged passengers had reduced survival advantages compared to young adults",
  "f_is_elderly", "Age_Groups", "Elderly passengers faced challenges in evacuation (mobility, strength, decision-making speed)",
  "f_age_class_deviation", "Age_Groups", "Deviation from class median age may indicate unusual passenger profile within class",
  
  # Fare-based features
  "f_fare_per_family", "Fare", "Higher fare per family member indicates greater wealth and better accommodations/proximity to lifeboats",
  "f_high_fare", "Fare", "Top quartile fare indicates first/second class passengers with premium lifeboat access",
  "f_low_fare", "Fare", "Bottom quartile fare indicates third class passengers with poor lifeboat access and location",
  "f_fare_class_ratio", "Fare", "Fare relative to class median identifies outlier prices within social class",
  
  # Class features
  "f_is_first_class", "Class", "First class indicator; strongest predictor due to lifeboat proximity and evacuation priority",
  "f_is_second_class", "Class", "Second class had moderate survival rates; better than third but worse than first",
  "f_is_third_class", "Class", "Third class indicator; lowest survival rates due to location and access restrictions",
  "f_is_upper_class", "Class", "Combined first and second class indicator; captures upper vs. lower class divide",
  
  # Interaction features
  "f_female_first_class", "Interaction", "Combination of highest priority factors: female gender + first class access",
  "f_female_second_class", "Interaction", "Female in second class; female priority minus one class disadvantage",
  "f_male_first_class", "Interaction", "Male in first class; gender disadvantage offset by first class advantage",
  "f_child_first_class", "Interaction", "Combination of two strong protective factors: childhood + first class",
  "f_child_female", "Interaction", "Child and female status; strongest protective combination in evacuation",
  "f_age_class_interaction", "Interaction", "Multiplicative effect: how age effects vary by class",
  "f_fare_class_interaction", "Interaction", "How fare premium varies in purchasing power effect across classes",
  
  # Cabin features
  "f_has_cabin_info", "Cabin", "Cabin information availability may correlate with passenger record quality and class status",
  "f_cabin_deck", "Cabin", "Cabin deck location determines proximity to lifeboats; decks A-C closer to lifeboats",
  
  # Ticket features
  "f_ticket_numeric_start", "Ticket", "Ticket number patterns may reflect booking method and passenger category",
  "f_ticket_length", "Ticket", "Ticket format variation may relate to passenger class or booking category",
  
  # Risk scores
  "f_female_score", "Risk_Score", "Composite score combining female gender advantage with class advantage",
  "f_third_class_risk", "Risk_Score", "Penalty score for third class; captures vulnerability without class benefits",
  "f_young_female_score", "Risk_Score", "Combined protective score: young age, female gender, and non-third class status",
  
  # Statistical features
  "f_age_zscore", "Statistical", "Standardized age allowing ML to interpret relative position in age distribution",
  "f_fare_zscore", "Statistical", "Standardized fare allowing comparison of relative spending within population"
)

# Display in sections for clarity
kable(filter(feature_rationale, Category %in% c("Original", "Embarked")), 
      caption = "Features 1-9: Original and Embarked Features")

kable(filter(feature_rationale, Category %in% c("Polynomial")), 
      caption = "Features 10-15: Polynomial and Transformation Features")

kable(filter(feature_rationale, Category %in% c("Family")), 
      caption = "Features 16-22: Family Structure Features")

kable(filter(feature_rationale, Category %in% c("Age_Groups")), 
      caption = "Features 23-27: Age-Based Categorical Features")

kable(filter(feature_rationale, Category %in% c("Fare")), 
      caption = "Features 28-31: Fare-Based Features")

kable(filter(feature_rationale, Category %in% c("Class")), 
      caption = "Features 32-35: Class Indicator Features")

kable(filter(feature_rationale, Category %in% c("Interaction")), 
      caption = "Features 36-42: Interaction Features")

kable(filter(feature_rationale, Category %in% c("Cabin", "Ticket")), 
      caption = "Features 43-45: Cabin and Ticket Features")

kable(filter(feature_rationale, Category %in% c("Risk_Score")), 
      caption = "Features 46-48: Composite Risk Scores")

kable(filter(feature_rationale, Category %in% c("Statistical")), 
      caption = "Features 49-50: Statistical Transformations")
```

Each of the `r n_features` features was designed with specific rationale. The features capture different dimensions of survival prediction: socio-economic factors (class, fare), biological factors (age, gender), family dynamics, and their interactions. The logical foundation for each feature derives from historical accounts of the evacuation and statistical patterns in survival data.

---

# 5. Data Preparation for Genetic Algorithm

```{r}
#| label: prepare-ga-data
#| eval: true

# Prepare data: separate features from target
X <- titanic_engineered %>%
  select(-target_survival) %>%
  as.matrix()

y <- titanic_engineered$target_survival %>%
  as.numeric()

# Create factor labels for caret (needs valid R variable names)
y_factor <- as.factor(ifelse(y == 1, "Yes", "No"))

# Check for any NaN or infinite values and replace with 0
X[is.na(X) | is.infinite(X)] <- 0

# Verify dimensions
dim(X)
length(y)
sum(y)  # Count of survivors
```

# The feature matrix contains `r nrow(X)` observations and `r ncol(X)` engineered features. The target variable has `r sum(y)` survivors out of `r length(y)` passengers, representing a `r round(100*sum(y)/length(y), 2)`% survival rate.

```{r}
#| label: define-ga-fitness
#| eval: true

# Define fitness function for genetic algorithm
# The fitness function trains a logistic regression model on selected features
# and returns the cross-validated AUC score as a measure of model performance

fitness_function <- function(features_binary) {
  # features_binary: binary vector indicating which features to include
  # If sum is 0, return worst possible fitness
  if (sum(features_binary) == 0) {
    return(-1)
  }
  
  # Select features based on binary indicator
  selected_features <- X[, which(features_binary == 1), drop = FALSE]
  
  # Handle case where all features are dropped
  if (ncol(selected_features) == 0) {
    return(-1)
  }
  
  # Train logistic regression using k-fold cross-validation
  tryCatch({
    # Create data frame for caret
    train_data <- data.frame(selected_features)
    train_data$target <- as.factor(ifelse(y == 1, "Yes", "No"))
    
    # Define cross-validation method
    train_control <- trainControl(
      method = "cv",
      number = 5,
      summaryFunction = twoClassSummary,
      classProbs = TRUE,
      returnResamp = "none"
    )
    
    # Train logistic regression model
    model <- train(
      target ~ .,
      data = train_data,
      method = "glm",
      family = "binomial",
      trControl = train_control,
      metric = "ROC"
    )
    
    # Return mean ROC as fitness value
    return(model$results$ROC)
  }, error = function(e) {
    return(-1)
  })
}

cat("Fitness function defined successfully\n")
```

We define a fitness function that evaluates the quality of a feature subset using 5-fold cross-validation logistic regression. The fitness metric is the mean Area Under the ROC Curve (AUC), which measures classification performance. This approach balances model accuracy with feature count (through evolutionary selection pressure).

---

# 6. Running the Genetic Algorithm

```{r}
#| label: run-genetic-algorithm
#| eval: true

# Run genetic algorithm for feature selection
# This may take some time as it requires training many logistic regression models

cat("Starting genetic algorithm for feature selection...\n")
cat("Number of features to select from:", ncol(X), "\n")
cat("Population size: 30, Generations: 20\n\n")

start_time <- Sys.time()

ga_result <- ga(
  type = "binary",  # Binary GA: 1 = include feature, 0 = exclude
  fitness = fitness_function,
  nBits = ncol(X),  # Number of features
  popSize = 30,  # Population size
  maxiter = 20,  # Number of generations
  run = 5,  # Stop if best fitness doesn't improve for 5 gens
  pmutation = 0.1,  # Mutation probability
  pcrossover = 0.8,  # Crossover probability
  parallel = FALSE,
  seed = 42,
  monitor = gaMonitor
)

end_time <- Sys.time()
runtime <- difftime(end_time, start_time, units = "mins")

cat("\n\nGenetic Algorithm Completed\n")
cat("Runtime:", round(runtime, 2), "minutes\n")
cat("Best fitness achieved:", round(ga_result@fitnessValue, 4), "\n")
cat("Number of generations run:", ga_result@iter, "\n")
```

The genetic algorithm executed for `r round(runtime, 2)` minutes, evaluating feature subsets across `r ga_result@iter` generations. The algorithm started with a population of 30 candidate solutions and applied evolutionary operators (selection, crossover, mutation) to progressively improve feature subsets based on their fitness (cross-validated AUC).

```{r}
#| label: extract-selected-features
#| eval: true

# Extract the best solution from the genetic algorithm
best_solution <- ga_result@solution[1, ]
n_selected_features <- sum(best_solution)

# Get names of selected features
selected_feature_names <- colnames(X)[which(best_solution == 1)]

# Create summary of selected features
selected_features_summary <- feature_rationale %>%
  filter(Feature_Name %in% selected_feature_names) %>%
  arrange(Feature_Name)

cat("Total features selected by GA:", n_selected_features, "\n")
cat("Reduction from original:", ncol(X), "→", n_selected_features, "\n")
cat("Feature reduction ratio:", round(100 * (1 - n_selected_features/ncol(X)), 2), "%\n\n")

kable(selected_features_summary, caption = "Features Selected by Genetic Algorithm")
```

The genetic algorithm selected `r n_selected_features` features from the original `r ncol(X)` candidates, achieving a `r round(100 * (1 - n_selected_features/ncol(X)), 2)`% reduction in feature dimensionality. This subset represents the algorithm's identification of features that collectively provide strong predictive signal for survival while minimizing redundancy.

---

# 7. Comparative Model Evaluation

```{r}
#| label: compare-all-features-model
#| eval: true

# Train model with ALL features
train_data_all <- data.frame(X)
train_data_all$target <- as.factor(ifelse(y == 1, "Yes", "No"))

train_control <- trainControl(
  method = "cv",
  number = 5,
  summaryFunction = twoClassSummary,
  classProbs = TRUE,
  returnResamp = "all"
)

model_all <- train(
  target ~ .,
  data = train_data_all,
  method = "glm",
  family = "binomial",
  trControl = train_control,
  metric = "ROC"
)

# Extract results
auc_all <- model_all$results$ROC
sensitivity_all <- model_all$results$Sens
specificity_all <- model_all$results$Spec

cat("All Features Model Performance (5-fold CV)\n")
cat("Mean AUC:", round(auc_all, 4), "\n")
cat("Mean Sensitivity:", round(sensitivity_all, 4), "\n")
cat("Mean Specificity:", round(specificity_all, 4), "\n")
```

The baseline model using all `r ncol(X)` features achieved a cross-validated AUC of `r round(auc_all, 4)`. This represents the predictive performance ceiling when using all available information without feature selection.

```{r}
#| label: compare-selected-features-model
#| eval: true

# Train model with SELECTED features only
selected_feature_indices <- which(best_solution == 1)
X_selected <- X[, selected_feature_indices, drop = FALSE]

train_data_selected <- data.frame(X_selected)
train_data_selected$target <- as.factor(ifelse(y == 1, "Yes", "No"))

model_selected <- train(
  target ~ .,
  data = train_data_selected,
  method = "glm",
  family = "binomial",
  trControl = train_control,
  metric = "ROC"
)

# Extract results
auc_selected <- model_selected$results$ROC
sensitivity_selected <- model_selected$results$Sens
specificity_selected <- model_selected$results$Spec

cat("Selected Features Model Performance (5-fold CV)\n")
cat("Mean AUC:", round(auc_selected, 4), "\n")
cat("Mean Sensitivity:", round(sensitivity_selected, 4), "\n")
cat("Mean Specificity:", round(specificity_selected, 4), "\n")
```

The feature-selected model using `r n_selected_features` features achieved a cross-validated AUC of `r round(auc_selected, 4)`. This demonstrates how the genetic algorithm identified a minimal but effective feature subset.

```{r}
#| label: model-comparison-visualization
#| eval: true

# Create comparison metrics
comparison_metrics <- tribble(
  ~Model, ~AUC, ~Sensitivity, ~Specificity, ~Num_Features,
  "All Features", auc_all, sensitivity_all, specificity_all, ncol(X),
  "GA Selected", auc_selected, sensitivity_selected, specificity_selected, n_selected_features
)

kable(comparison_metrics, caption = "Model Performance Comparison")

# Visualization of metric comparison
comparison_long <- comparison_metrics %>%
  pivot_longer(
    cols = c(AUC, Sensitivity, Specificity),
    names_to = "Metric",
    values_to = "Score"
  )

ggplot(comparison_long, aes(x = Metric, y = Score, fill = Model)) +
  geom_col(position = "dodge", width = 0.7) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    legend.position = "bottom",
    axis.title.x = element_blank()
  ) +
  labs(
    title = "Model Performance: All Features vs. GA-Selected Features",
    y = "Score (0-1)"
  ) +
  ylim(0, 1) +
  scale_fill_brewer(palette = "Set2")
```

The comparison visualization shows that the GA-selected feature subset maintains comparable performance to the full model while using `r round(100 * n_selected_features/ncol(X), 1)`% of the original feature count. This demonstrates the efficiency gain from feature selection.

---

# 8. Feature Importance in Selected Subset

```{r}
#| label: selected-model-coefficients
#| eval: true

# Extract coefficients from the selected features model
# Note: The final model is trained on the entire training set for interpretation

final_model <- glm(
  as.formula(paste("target ~ ", paste(colnames(X_selected), collapse = " + "))),
  data = train_data_selected,
  family = "binomial"
)

# Extract and organize coefficients
coefficients_df <- tibble(
  Feature = names(coef(final_model))[-1],  # Exclude intercept
  Coefficient = coef(final_model)[-1],
  Abs_Coefficient = abs(coef(final_model)[-1])
) %>%
  arrange(desc(Abs_Coefficient))

kable(head(coefficients_df, 10), caption = "Top 10 Most Important Features (by Absolute Coefficient)")

# Visualize top coefficients
top_features <- head(coefficients_df, 10) %>%
  mutate(Feature = fct_reorder(Feature, Coefficient))

ggplot(top_features, aes(x = Feature, y = Coefficient, fill = Coefficient > 0)) +
  geom_col() +
  coord_flip() +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    legend.position = "bottom",
    axis.title.y = element_blank()
  ) +
  labs(
    title = "Top 10 Feature Coefficients in Selected Model",
    y = "Logistic Regression Coefficient",
    fill = "Positive Effect"
  ) +
  scale_fill_manual(values = c("FALSE" = "#d7191c", "TRUE" = "#2b83ba"))
```

The logistic regression coefficients on the GA-selected features reveal the direction and magnitude of each feature's influence on survival prediction. Positive coefficients increase survival probability, while negative coefficients decrease it. The top features selected by the GA show clear interpretability aligned with historical survival patterns.

---

# 9. Genetic Algorithm Evolution Summary

```{r}
#| label: ga-evolution-analysis
#| eval: true

# Create summary of GA execution
ga_summary <- tribble(
  ~Metric, ~Value,
  "Algorithm Type", "Binary Genetic Algorithm",
  "Population Size", "30",
  "Number of Generations", as.character(ga_result@iter),
  "Mutation Probability", "0.1",
  "Crossover Probability", "0.8",
  "Fitness Metric", "5-fold CV AUC",
  "Initial Features", as.character(ncol(X)),
  "Selected Features", as.character(n_selected_features),
  "Best Fitness Score", as.character(round(ga_result@fitnessValue, 4)),
  "Runtime (minutes)", as.character(round(runtime, 2))
)

kable(ga_summary, caption = "Genetic Algorithm Execution Summary")

# Calculate performance gains
auc_improvement <- ((auc_selected - auc_all) / auc_all) * 100
feature_reduction <- ((ncol(X) - n_selected_features) / ncol(X)) * 100

cat("\nKey Insights:\n")
cat("- Feature reduction: ", round(feature_reduction, 1), "%\n", sep = "")
cat("- AUC change: ", round(auc_improvement, 2), "%\n", sep = "")
cat("- Features per observation ratio: All=", round(ncol(X)/nrow(X), 3), 
    " → Selected=", round(n_selected_features/nrow(X), 3), "\n", sep = "")
```

The genetic algorithm successfully identified a reduced feature set maintaining model performance. The `r round(feature_reduction, 1)`% reduction in features improves computational efficiency and model interpretability while preserving the AUC at `r round(auc_selected, 4)`.

---

# 10. Conclusion and Next Steps

## Summary of Analysis

This analysis demonstrated a complete workflow for genetic algorithm-based feature selection on the Titanic survival prediction task:

1. **Feature Engineering**: Created 50 candidate features through mathematical transformations, domain interactions, and statistical scaling
2. **Genetic Algorithm**: Applied GA to search through feature combinations, identifying optimal subsets based on 5-fold CV AUC
3. **Model Evaluation**: Compared full model (AUC = `r round(auc_all, 4)`) with GA-selected model (AUC = `r round(auc_selected, 4)`)
4. **Feature Reduction**: Achieved `r round(feature_reduction, 1)`% reduction from `r ncol(X)` to `r n_selected_features` features

## Interpretation Framework

The genetic algorithm's selection of `r n_selected_features` features represents features that:
- Provide complementary information (low redundancy)
- Have strong individual signal for survival prediction
- Interact effectively with other selected features
- Maintain generalization performance in cross-validation

## Practical Implications

The selected feature subset enables:
- **Reduced computational cost**: Fewer features mean faster training and inference
- **Improved interpretability**: Simpler models with fewer parameters are easier to explain
- **Better generalization**: Feature selection can reduce overfitting and improve test set performance
- **Domain alignment**: Selected features align with historical factors known to affect survival

## Limitations and Considerations

- The genetic algorithm found a near-optimal solution but not necessarily the global optimum
- Cross-validation results may vary based on fold assignment
- Feature selection is dataset-specific; generalization to new data should be validated
- The 50-feature space was engineer-defined; domain expertise guided feature creation

---

# 11. Session Information

```{r}
#| label: session-info
#| eval: true

sessionInfo()
```
