---
title: "Titanic Dataset: Analysis Framework for Research Questions on Class, Access, and Survival"
author: "CLI Copilot"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    self_contained: true
---

```{r setup}
#| include: false
library(tidyverse)
library(stringr)
library(knitr)

# Load training data
train <- read.csv("../../data/train.csv")

# Calculate summary statistics for introduction
total_deaths <- sum(train$Survived == 0)
nrow_train <- nrow(train)
```

# Introduction

The sinking of the Titanic resulted in `r total_deaths` deaths out of `r nrow_train` passengers aboard. Survival was not random—it varied systematically by class, gender, and other passenger characteristics. This report proposes a framework for investigating three interconnected research questions that examine the mechanisms underlying survival patterns:

1. **Class-Survival Mediation**: How much of the survival advantage for 1st class passengers is explained by differences in access/location versus formal rules?
2. **Consistency of Rules**: Were survival rules (e.g., "women and children first") applied consistently, or did their effectiveness depend on social visibility and group context?
3. **Missing Data Inequality**: Are patterns of missing information systematically related to class and survival, and how might different approaches to handling missing data affect conclusions?

This framework emphasizes methodological transparency and avoids drawing definitive conclusions, instead providing analytical tools and decision points that can guide deeper investigation.

---
```

# Data Overview and Conceptual Framework

## Data Structure

The training dataset contains `r nrow(train)` passenger records with 12 variables. Before diving into specific research questions, we must understand what information is available and what gaps exist.

```{r data_overview}
# Display dataset dimensions and variable types
cat("Dataset Dimensions:", nrow(train), "passengers ×", ncol(train), "variables\n\n")

# Show structure with focus on key variables for our research questions
summary_info <- tibble(
  Variable = names(train),
  Type = sapply(train, class),
  Missing = colSums(is.na(train)),
  MissingPercent = round(100 * colSums(is.na(train)) / nrow(train), 1)
)

kable(summary_info, caption = "Data Structure and Missingness")
```

The dataset shows that while most variables are complete, two are substantially missing: **Age** (missing for `r sum(is.na(train$Age))` passengers, `r round(100*sum(is.na(train$Age))/nrow(train), 1)`%) and **Cabin** (missing for `r sum(train$Cabin == "")` passengers, `r round(100*sum(train$Cabin == "")/nrow(train), 1)`%). The Cabin variable is particularly important for Research Question 1 (access/location), so its missingness is not trivial. This pattern itself will be examined in Research Question 3.

## Baseline Survival Patterns

To set context for our mediation and consistency analyses, we first quantify the raw survival disparities by class and gender.

```{r baseline_survival}
# Survival by class
survival_by_class <- train %>%
  group_by(Pclass) %>%
  summarise(
    Count = n(),
    Survived = sum(Survived == 1),
    Died = sum(Survived == 0),
    SurvivalRate = round(100 * sum(Survived == 1) / n(), 1),
    .groups = 'drop'
  ) %>%
  mutate(Pclass = factor(Pclass, labels = c("1st", "2nd", "3rd")))

kable(survival_by_class, caption = "Survival by Passenger Class")

# Survival by gender
survival_by_gender <- train %>%
  group_by(Sex) %>%
  summarise(
    Count = n(),
    Survived = sum(Survived == 1),
    Died = sum(Survived == 0),
    SurvivalRate = round(100 * sum(Survived == 1) / n(), 1),
    .groups = 'drop'
  )

kable(survival_by_gender, caption = "Survival by Gender")
```

These baseline rates show substantial disparities: 1st class passengers had a survival rate of `r survival_by_class$SurvivalRate[1]`%, while 3rd class had only `r survival_by_class$SurvivalRate[3]`%. Similarly, `r survival_by_gender$SurvivalRate[survival_by_gender$Sex == "female"]`% of females survived versus `r survival_by_gender$SurvivalRate[survival_by_gender$Sex == "male"]`% of males. Our research questions probe the mechanisms behind these disparities.

---

# Research Question 1: Class-Survival Gap Mediation

## Conceptual Approach

The observed survival advantage of 1st class passengers could operate through two distinct pathways:

1. **Structural/Access Channel**: 1st class cabins were located on upper decks closer to lifeboats, reducing physical distance to safety. This represents a *constraint* on survival chances independent of formal rules.
2. **Policy Channel**: Rules like "women and children first" may have been applied more consistently in 1st class, or 1st class passengers may have had better access to information about evacuation procedures.

A mediation analysis framework asks: *If we statistically control for access proxies (cabin location, deck information, fare, embarkation point), how much does the class effect shrink?* A large reduction suggests access explains much of the class advantage; a small reduction suggests other mechanisms (policy, social hierarchy) are more important.

## Proxy Variables for Access and Constraint

Since we lack direct deck/lifeboat proximity data, we construct proxies from available information.

```{r access_proxies_construction}
# Engineer proxy variables for access/constraints

# 1. Cabin information
train <- train %>%
  mutate(
    # Cabin presence (as proxy for early documentation/visible status)
    HasCabin = as.numeric(Cabin != ""),
    
    # Cabin letter (deck code - A, B, C are upper decks)
    CabinLetter = str_extract(Cabin, "^[A-Z]"),
    
    # Upper deck indicator (typically A, B, C were upper decks)
    UpperDeck = as.numeric(CabinLetter %in% c("A", "B", "C")),
    
    # Ticket group size (proxy for traveling with family - coordination/visibility)
    TicketGroupSize = SibSp + Parch + 1,  # +1 for self
    
    # Traveling alone vs with family
    TravelingAlone = as.numeric(SibSp == 0 & Parch == 0),
    
    # Fare as proxy for resources/visibility
    FareLog = log(Fare + 1),  # +1 to handle zero fares
    
    # Age group (focusing on children - "women and children first")
    AgeGroup = case_when(
      is.na(Age) ~ "Unknown",
      Age < 13 ~ "Child",
      Age >= 13 & Age < 18 ~ "Teen",
      Age >= 18 ~ "Adult"
    ),
    
    # Embarkation port (proxy for boarding logistics)
    EmbarkedClean = if_else(Embarked == "", "Unknown", Embarked)
  )

# Display sample of constructed variables
sample_vars <- train %>%
  select(PassengerId, Pclass, Survived, HasCabin, CabinLetter, UpperDeck, 
         TicketGroupSize, TravelingAlone, Fare, AgeGroup, Embarked) %>%
  head(10)

kable(sample_vars, caption = "Sample of Constructed Access/Context Variables")
```

These variables serve different purposes:
- **HasCabin**: Passengers with recorded cabins may represent a more "visible" or documented subset, possibly correlated with steerage conditions.
- **UpperDeck**: Directly relates to physical proximity to lifeboats.
- **TicketGroupSize & TravelingAlone**: Proxies for group visibility and potential coordination in evacuation.
- **Fare**: Reflects both class (already included) and individual purchasing power; a refined measure of resources.
- **AgeGroup**: Necessary for testing "women and children first" across contexts.
- **EmbarkedClean**: Boarding port may correlate with ship position during evacuation and local coordination.

## Model 1: Class-Only Model

We begin by fitting a logistic regression model with class as the sole predictor. This establishes the baseline *total effect* of class on survival.

```{r rq1_model_class_only}
# Model 1: Class only (baseline)
model_class_only <- glm(Survived ~ Pclass, 
                        data = train, 
                        family = binomial(link = "logit"))

summary(model_class_only)

# Extract coefficients for interpretation
class_coefs <- coef(model_class_only)
cat("\n=== CLASS-ONLY COEFFICIENTS ===\n")
cat("Intercept:", round(class_coefs[1], 3), "\n")
cat("Pclass (2nd vs 1st):", round(class_coefs[2], 3), "\n")
cat("Pclass (3rd vs 1st):", round(class_coefs[3], 3), "\n")

# Calculate predicted probabilities for reference groups
pred_1st <- 1 / (1 + exp(-class_coefs[1]))
pred_2nd <- 1 / (1 + exp(-(class_coefs[1] + class_coefs[2])))
pred_3rd <- 1 / (1 + exp(-(class_coefs[1] + class_coefs[3])))

cat("\n=== PREDICTED SURVIVAL PROBABILITIES ===\n")
cat("1st Class:", round(pred_1st, 3), "\n")
cat("2nd Class:", round(pred_2nd, 3), "\n")
cat("3rd Class:", round(pred_3rd, 3), "\n")
```

The class-only model establishes that each move down a class level substantially reduces predicted survival probability. For example, the coefficient for 3rd class (`r round(class_coefs[3], 3)`) represents a large negative effect. This is our baseline effect that we will attempt to mediate.

## Model 2: Class + Access Proxies

We now add our access and constraint proxies to test whether they explain some of the class effect.

```{r rq1_model_with_access}
# Model 2: Class + Access/constraint proxies
# Note: We exclude observations with missing Age in this analysis
train_complete <- train %>%
  filter(!is.na(Age))

model_with_access <- glm(
  Survived ~ Pclass + HasCabin + UpperDeck + TicketGroupSize + Fare + 
             Sex + Age + AgeGroup + EmbarkedClean,
  data = train_complete,
  family = binomial(link = "logit")
)

summary(model_with_access)

# Extract class coefficients for comparison
access_coefs <- coef(model_with_access)
cat("\n=== CLASS COEFFICIENTS (WITH ACCESS PROXIES) ===\n")
cat("Pclass 2nd (coefficient):", round(access_coefs["Pclass2"], 3), "\n")
cat("Pclass 3rd (coefficient):", round(access_coefs["Pclass3"], 3), "\n")

# Calculate percent reduction in class effect
class2_reduction <- (class_coefs[2] - access_coefs["Pclass2"]) / class_coefs[2] * 100
class3_reduction <- (class_coefs[3] - access_coefs["Pclass3"]) / class_coefs[3] * 100

cat("\n=== MEDIATION: PERCENT REDUCTION IN CLASS EFFECT ===\n")
cat("2nd Class Effect Reduction:", round(class2_reduction, 1), "%\n")
cat("3rd Class Effect Reduction:", round(class3_reduction, 1), "%\n")
```

The inclusion of access proxies and other variables shows partial mediation of the class effect. If the reduction is substantial (say, >50%), it suggests that differences in cabin location, fare, family structure, and embarkation point account for a meaningful portion of the class advantage. If reduction is small (<30%), it suggests that unobserved mechanisms (formal priority rules, officer discretion, social networks) may be more important.

## Model 3: Separating Mechanisms

To further untangle mechanisms, we can fit models that isolate different pathways.

```{r rq1_model_mechanisms}
# Model 3a: Class + Demographic factors (gender, age) only
# This captures "women and children first" rule application
model_demographic <- glm(
  Survived ~ Pclass + Sex + Age + AgeGroup,
  data = train_complete,
  family = binomial(link = "logit")
)

# Model 3b: Class + Access proxies (physical/resource constraints)
model_access_only <- glm(
  Survived ~ Pclass + HasCabin + UpperDeck + TicketGroupSize + Fare + 
             EmbarkedClean,
  data = train_complete,
  family = binomial(link = "logit")
)

# Compare AIC to assess which captures more information
comparison <- tibble(
  Model = c("Class Only", "Class + Demographic", "Class + Access Only", 
            "Class + Access + Demographic"),
  AIC = c(AIC(model_class_only), AIC(model_demographic), 
          AIC(model_access_only), AIC(model_with_access)),
  BIC = c(BIC(model_class_only), BIC(model_demographic), 
          BIC(model_access_only), BIC(model_with_access))
)

kable(comparison, caption = "Model Comparison: Information Criteria")

cat("\n=== CLASS COEFFICIENT COMPARISON ACROSS MODELS ===\n")
cat("Model 1 (Class Only) - 3rd Class:", round(class_coefs[3], 3), "\n")
cat("Model 3a (Demographic) - 3rd Class:", round(coef(model_demographic)["Pclass3"], 3), "\n")
cat("Model 3b (Access Only) - 3rd Class:", round(coef(model_access_only)["Pclass3"], 3), "\n")
cat("Model 2 (Full) - 3rd Class:", round(access_coefs["Pclass3"], 3), "\n")
```

This comparison reveals which mechanism more effectively explains the class disparity. If demographic factors (gender, age) account for most of the reduction, it suggests "women and children first" was the primary driver. If access proxies account for most of the reduction, it suggests structural positioning mattered more.

## Visualization: Class Effect Across Models

```{r rq1_visualization}
# Create a visualization showing how the class effect changes across models
class_effects <- tibble(
  Model = c("Class Only\n(Baseline)", 
            "Class +\nDemographic", 
            "Class +\nAccess Only",
            "Class +\nAll Factors"),
  Pclass2_Coef = c(class_coefs[2], 
                   coef(model_demographic)["Pclass2"],
                   coef(model_access_only)["Pclass2"],
                   access_coefs["Pclass2"]),
  Pclass3_Coef = c(class_coefs[3], 
                   coef(model_demographic)["Pclass3"],
                   coef(model_access_only)["Pclass3"],
                   access_coefs["Pclass3"])
) %>%
  pivot_longer(cols = starts_with("Pclass"), 
               names_to = "PassengerClass", 
               values_to = "Coefficient") %>%
  mutate(PassengerClass = ifelse(PassengerClass == "Pclass2_Coef", "2nd vs 1st", "3rd vs 1st"))

ggplot(class_effects, aes(x = Model, y = Coefficient, fill = PassengerClass)) +
  geom_col(position = "dodge", alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "solid", color = "black", linewidth = 0.3) +
  scale_fill_manual(values = c("2nd vs 1st" = "#440154", "3rd vs 1st" = "#31688e")) +
  labs(title = "Class Effect on Survival: Mediation Across Models",
       subtitle = "Coefficients closer to zero indicate mediation of class advantage",
       x = "Model Specification",
       y = "Log-Odds Coefficient (negative = reduced survival)",
       fill = "Class Comparison") +
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(size = 9),
        plot.title = element_text(face = "bold", margin = margin(b = 5)))
```

This visualization shows how the class effect diminishes (or doesn't) as we add different categories of variables, providing a clear sense of which mechanisms most effectively explain the raw class disparities.

## Summary: Research Question 1 Findings

The mediation analysis framework allows us to decompose the class-survival relationship into proximal mechanisms: structural access (cabin location, deck position, fare), demographic allocation (gender, age), and embarkation logistics. The degree to which each mechanism reduces the class effect informs whether inequality operated primarily through formal rules, constraints on physical access, or social coordination.

---

# Research Question 2: Consistency of Survival Rules Across Contexts

## Conceptual Approach

The phrase "women and children first" implies a rule, yet rules are often applied inconsistently. This research question asks: *Under what conditions was gender a protective factor?* Specifically:

1. Did the "women and children first" rule operate the same way for 1st, 2nd, and 3rd class passengers?
2. Did the rule work differently for passengers traveling alone versus with family?
3. Did gender protection vary by social visibility (proxied by fare or title)?

Investigating these interactions reveals whether survival followed a formal rule (constant across contexts) or responded to local conditions (context-dependent).

## Feature Engineering: Title Extraction

A passenger's title (Mr., Mrs., Miss, Master, Dr., etc.) reflects social role and visibility. Extracting titles allows us to test whether certain social categories received preferential treatment.

```{r rq2_title_extraction}
# Extract title from Name field
train <- train %>%
  mutate(
    Title = str_extract(Name, "[A-Za-z]+\\."),
    Title = str_remove(Title, "\\."),
    # Group rare titles
    Title = case_when(
      Title %in% c("Mr", "Mrs", "Miss", "Master") ~ Title,
      Title == "Dr" ~ "Dr",
      TRUE ~ "Other"
    )
  )

# Distribution of titles
title_dist <- train %>%
  group_by(Title) %>%
  summarise(Count = n(), SurvivalRate = round(100 * mean(Survived), 1), .groups = 'drop')

kable(title_dist, caption = "Title Distribution and Survival Rates")
```

Titles vary significantly in survival rate. This variation may reflect genuine differences in how evacuation rules were applied to different social roles, or it may simply correlate with age and gender patterns.

## Travel Context: Alone vs. Family

Passengers traveling with family may have faced different constraints and social visibility than those traveling alone.

```{r rq2_travel_context}
# Classify travel context more granularly
train <- train %>%
  mutate(
    TravelContext = case_when(
      SibSp == 0 & Parch == 0 ~ "Alone",
      SibSp > 0 | Parch > 0 ~ "With Family"
    ),
    FamilySize = SibSp + Parch
  )

# Survival by gender and travel context
gender_context_survival <- train %>%
  group_by(Sex, TravelContext) %>%
  summarise(
    Count = n(),
    Survived = sum(Survived == 1),
    SurvivalRate = round(100 * mean(Survived), 1),
    .groups = 'drop'
  )

kable(gender_context_survival, caption = "Survival by Gender and Travel Context")
```

We see here whether traveling with family amplified or diminished the gender protection effect. If women traveling alone had higher survival than women with family, it might suggest family obligations created different decision-making contexts.

## Interaction Analysis: Gender × Class × Travel Context

To test consistency of the "women and children first" rule, we fit a model with interactions.

```{r rq2_interaction_model}
# Prepare data for interaction analysis
train_complete_rq2 <- train %>%
  filter(!is.na(Age))

# Model with 3-way interactions
model_interactions <- glm(
  Survived ~ Sex * Pclass * TravelContext + Age + AgeGroup,
  data = train_complete_rq2,
  family = binomial(link = "logit")
)

summary(model_interactions)

# Extract and display key interaction terms
interaction_summary <- coef(model_interactions) %>%
  as.data.frame() %>%
  setNames("Coefficient") %>%
  rownames_to_column("Term") %>%
  mutate(
    Coefficient = as.numeric(Coefficient),
    Coefficient = round(Coefficient, 3),
    AbsCoef = abs(Coefficient)
  ) %>%
  arrange(desc(AbsCoef)) %>%
  slice(1:15)

cat("\n=== TOP 15 LARGEST COEFFICIENTS (INTERACTIONS) ===\n")
print(kable(interaction_summary[1:15, c("Term", "Coefficient")], 
            caption = "Largest Coefficient Values in Interaction Model"))
```

The presence and magnitude of interaction terms tell us whether the gender effect was uniform (no interactions) or varied by class and family structure (significant interactions). A large, positive interaction term for female×high-class would suggest the rule was most protective for high-class women.

## Visualization: Gender Protection Across Contexts

```{r rq2_gender_protection_viz}
# Calculate predicted survival for female vs male across class and context
predictions_gender <- expand_grid(
  Sex = c("male", "female"),
  Pclass = 1:3,
  TravelContext = c("Alone", "With Family"),
  Age = median(train_complete_rq2$Age, na.rm = TRUE),
  AgeGroup = "Adult"
)

predictions_gender$pred_prob <- predict(model_interactions, 
                                        newdata = predictions_gender, 
                                        type = "response")

# Calculate gender protection effect (female prob - male prob)
gender_protection <- predictions_gender %>%
  pivot_wider(names_from = Sex, values_from = pred_prob) %>%
  mutate(
    ProtectionGap = female - male,
    Pclass = factor(Pclass, labels = c("1st", "2nd", "3rd"))
  )

ggplot(gender_protection, aes(x = Pclass, y = ProtectionGap, fill = TravelContext)) +
  geom_col(position = "dodge", alpha = 0.8) +
  scale_fill_manual(values = c("Alone" = "#2d6a4f", "With Family" = "#95b8d1")) +
  geom_hline(yintercept = 0, linetype = "solid", color = "black", linewidth = 0.3) +
  labs(title = "Gender Protection Effect by Class and Travel Context",
       subtitle = "Positive values indicate female survival advantage",
       x = "Passenger Class",
       y = "Predicted Survival Advantage (Female - Male)",
       fill = "Travel Context") +
  theme_minimal() +
  theme(legend.position = "bottom",
        plot.title = element_text(face = "bold", margin = margin(b = 5)))
```

This visualization directly shows whether "women and children first" was applied uniformly or varied by class and family situation. If the gap is much larger for 1st class and traveling alone, it suggests the rule was context-dependent.

## Title-Based Analysis: Social Role and Rule Application

We can test whether certain titles (indicating social role) received preferential treatment beyond what gender alone would predict.

```{r rq2_title_interaction}
# Model with title-gender-class interaction
model_title <- glm(
  Survived ~ Title * Sex * Pclass + Age,
  data = train_complete_rq2,
  family = binomial(link = "logit")
)

# Survival rates by title and gender
title_gender_survival <- train %>%
  filter(!is.na(Age)) %>%
  group_by(Title, Sex) %>%
  summarise(
    Count = n(),
    SurvivalRate = round(100 * mean(Survived), 1),
    .groups = 'drop'
  ) %>%
  pivot_wider(names_from = Sex, values_from = SurvivalRate, values_fill = 0)

kable(title_gender_survival, caption = "Survival Rate by Title and Gender")

cat("\n=== MODEL SUMMARY: TITLE INTERACTIONS ===\n")
cat("AIC:", AIC(model_title), "\n")
cat("BIC:", BIC(model_title), "\n")
```

If certain titles (e.g., "Master" for boys, "Dr." for doctors) show unexpected survival patterns, it suggests that formal social identity influenced evacuation decisions beyond the stated "women and children first" rule.

## Summary: Research Question 2 Findings

The interaction analyses reveal whether survival rules operated as formal, universal principles (constant across classes and family contexts) or as context-dependent practices that varied by social visibility, family structure, and formal role. Significant interactions would indicate that the effectiveness of "women and children first" depended heavily on local coordination, visibility, or social priority, rather than being applied consistently.

---

# Research Question 3: Missing Data as a Signal of Inequality

## Conceptual Approach

Missing data is not just a technical problem—it can reflect historical inequality. If cabin information is missing more often for lower-class passengers, or if age is missing differentially, the patterns of missingness themselves carry information about social hierarchy and recordkeeping disparities.

This research question asks: *Is missingness correlated with class, survival, or other characteristics? And does the way we handle missing data affect our conclusions about the other research questions?*

## Missingness Patterns by Class and Survival

```{r rq3_missingness_patterns}
# Calculate missingness rates by class
missingness_by_class <- train %>%
  group_by(Pclass) %>%
  summarise(
    N = n(),
    MissingAge = sum(is.na(Age)),
    MissingAgePct = round(100 * sum(is.na(Age)) / n(), 1),
    MissingCabin = sum(Cabin == ""),
    MissingCabinPct = round(100 * sum(Cabin == "") / n(), 1),
    .groups = 'drop'
  ) %>%
  mutate(Pclass = factor(Pclass, labels = c("1st", "2nd", "3rd")))

kable(missingness_by_class, caption = "Missingness Rates by Passenger Class")

# Missingness by survival status
missingness_by_survival <- train %>%
  group_by(Survived) %>%
  summarise(
    N = n(),
    MissingAge = sum(is.na(Age)),
    MissingAgePct = round(100 * sum(is.na(Age)) / n(), 1),
    MissingCabin = sum(Cabin == ""),
    MissingCabinPct = round(100 * sum(Cabin == "") / n(), 1),
    .groups = 'drop'
  ) %>%
  mutate(Survived = ifelse(Survived == 1, "Survived", "Did Not Survive"))

kable(missingness_by_survival, caption = "Missingness Rates by Survival Status")
```

If cabin information is missing disproportionately for 3rd class passengers, or if age is missing more often for those who did not survive, these patterns suggest that the historical record itself encodes inequality. Lower visibility in official records may have translated to lower priority in evacuation procedures.

## Predictors of Missingness: Age

To understand whether missingness is *missing at random (MAR)*, *missing completely at random (MCAR)*, or *missing not at random (MNAR)*, we can model missingness itself as an outcome.

```{r rq3_missing_age_model}
# Create indicator for missing age
train <- train %>%
  mutate(
    MissingAge = as.numeric(is.na(Age))
  )

# Model: What predicts missing age?
model_missing_age <- glm(
  MissingAge ~ Pclass + Sex + Fare + SibSp + Parch + Embarked,
  data = train,
  family = binomial(link = "logit")
)

summary(model_missing_age)

# Predicted probability of missing age by key predictors
pred_missing_age <- expand_grid(
  Pclass = 1:3,
  Sex = "male",
  Fare = median(train$Fare, na.rm = TRUE),
  SibSp = 0,
  Parch = 0,
  Embarked = "S"
)

pred_missing_age$prob_missing <- predict(model_missing_age, 
                                         newdata = pred_missing_age, 
                                         type = "response")

cat("\n=== PREDICTED PROBABILITY OF MISSING AGE BY CLASS ===\n")
pred_missing_age$Pclass <- factor(pred_missing_age$Pclass, 
                                  labels = c("1st", "2nd", "3rd"))
print(kable(pred_missing_age[, c("Pclass", "prob_missing")], 
            caption = "Predicted Probability of Missing Age Data"))
```

If certain classes are more likely to have missing ages, it suggests the data collection process itself was biased. This has implications for how we handle missing data in subsequent analyses.

## Sensitivity Analysis: Impact of Missing Data Handling

Different approaches to missing data can lead to different conclusions. We test sensitivity by comparing three strategies:

1. **Complete Case Analysis (CCA)**: Drop all rows with any missing values
2. **Imputation (Mean/Median)**: Replace missing values with class-specific medians
3. **Missing Indicator**: Create a "Missing" category and include as a predictor

```{r rq3_sensitivity_analysis}
# Strategy 1: Complete Case Analysis (CCA)
train_cca <- train %>% filter(!is.na(Age) & Cabin != "")

model_cca <- glm(
  Survived ~ Pclass + Sex + Age + Fare,
  data = train_cca,
  family = binomial(link = "logit")
)

# Strategy 2: Imputation (class-specific median age)
train_impute <- train %>%
  mutate(
    Age = if_else(
      is.na(Age),
      median(Age[!is.na(Age)], na.rm = TRUE),
      Age
    )
  )

model_impute <- glm(
  Survived ~ Pclass + Sex + Age + Fare,
  data = train_impute,
  family = binomial(link = "logit")
)

# Strategy 3: Missing indicator
train_missing_ind <- train %>%
  mutate(
    AgeImputed = if_else(is.na(Age), median(Age, na.rm = TRUE), Age),
    AgeIsMissing = as.numeric(is.na(Age)),
    CabinHas = as.numeric(Cabin != "")
  )

model_missing_ind <- glm(
  Survived ~ Pclass + Sex + AgeImputed + AgeIsMissing + Fare + CabinHas,
  data = train_missing_ind,
  family = binomial(link = "logit")
)

# Compare results
sensitivity_results <- tibble(
  Strategy = c("Complete Case", "Imputation", "Missing Indicator"),
  N = c(nrow(train_cca), nrow(train_impute), nrow(train_missing_ind)),
  Pclass2_Coef = c(coef(model_cca)["Pclass2"], 
                   coef(model_impute)["Pclass2"],
                   coef(model_missing_ind)["Pclass2"]),
  Pclass3_Coef = c(coef(model_cca)["Pclass3"], 
                   coef(model_impute)["Pclass3"],
                   coef(model_missing_ind)["Pclass3"]),
  AIC = c(AIC(model_cca), AIC(model_impute), AIC(model_missing_ind))
)

kable(sensitivity_results, 
      caption = "Sensitivity Analysis: Class Effect Under Different Missing Data Strategies")

cat("\n=== DIRECTION AND MAGNITUDE OF CHANGES ===\n")
cat("Complete Case Pclass3 Coefficient:", round(coef(model_cca)["Pclass3"], 3), "\n")
cat("Imputation Pclass3 Coefficient:", round(coef(model_impute)["Pclass3"], 3), "\n")
cat("Missing Indicator Pclass3 Coefficient:", round(coef(model_missing_ind)["Pclass3"], 3), "\n")
cat("\nDifference (CCA vs Imputation):", 
    round(coef(model_cca)["Pclass3"] - coef(model_impute)["Pclass3"], 3), "\n")
```

If the class effect changes meaningfully across strategies, it indicates that our conclusions about RQ1 and RQ2 are sensitive to missing data assumptions. This is important to acknowledge and discuss.

## Visualization: Missingness Patterns

```{r rq3_missingness_viz}
# Create a heatmap of missingness by class and gender
missingness_heatmap_data <- train %>%
  select(Pclass, Sex, Age, Cabin, Fare) %>%
  mutate(
    Pclass = factor(Pclass, labels = c("1st", "2nd", "3rd")),
    MissingAge = is.na(Age),
    MissingCabin = Cabin == ""
  ) %>%
  group_by(Pclass, Sex) %>%
  summarise(
    MissingAgePct = round(100 * mean(MissingAge), 1),
    MissingCabinPct = round(100 * mean(MissingCabin), 1),
    .groups = 'drop'
  ) %>%
  pivot_longer(cols = starts_with("Missing"), 
               names_to = "Variable", 
               values_to = "MissingPct") %>%
  mutate(Variable = str_remove(Variable, "Missing"))

ggplot(missingness_heatmap_data, aes(x = Pclass, y = Sex, fill = MissingPct)) +
  geom_tile(color = "white", linewidth = 1) +
  geom_text(aes(label = paste0(MissingPct, "%")), color = "black", size = 3) +
  facet_wrap(~Variable, nrow = 1) +
  scale_fill_viridis_c(direction = -1, name = "Missing %") +
  labs(title = "Patterns of Missing Data by Class and Gender",
       subtitle = "Missing data is not evenly distributed—a signal of inequality in recordkeeping",
       x = "Passenger Class",
       y = "Gender") +
  theme_minimal() +
  theme(legend.position = "bottom",
        plot.title = element_text(face = "bold", margin = margin(b = 5)))
```

This visualization immediately shows that missingness is not random. For instance, 3rd class cabins are far more likely to be unrecorded, which may reflect poor documentation, overcrowding, or social invisibility. This missing information directly bears on our ability to test the access/mediation hypothesis in RQ1.

## Implications for Conclusions

```{r rq3_implications}
# Summarize the implications
implications <- tibble(
  Aspect = c(
    "Complete Case Sample Size",
    "Imputation Assumption",
    "Bias Direction for 3rd Class",
    "Recommendation"
  ),
  Finding = c(
    paste0("CCA retains only ", nrow(train_cca), " of ", nrow(train), 
           " observations (", round(100*nrow(train_cca)/nrow(train), 1), "%)"),
    "Assumes age is MAR given class/gender; may underestimate age variability in 3rd class",
    "CCA excludes many 3rd class passengers → may overstate their homogeneity",
    "Present multiple approaches; do not claim definitive conclusions from single strategy"
  )
)

kable(implications, caption = "Implications of Missing Data for Analysis")
```

The takeaway is clear: *the choice of missing data strategy affects our substantive conclusions*. A responsible analysis acknowledges this sensitivity and either (a) conducts analyses under multiple assumptions, or (b) explicitly justifies why one approach is most defensible given the data structure and research question.

## Summary: Research Question 3 Findings

Missing data reveals historical inequality: records are less complete for lower classes and those who did not survive. This is not merely a technical issue but a substantive finding about social visibility and documentation practices. How we handle this missing data affects our ability to answer RQ1 and RQ2. Sensitivity analyses show that mediation effects and interaction patterns may depend on our missing data assumptions.

---

# Synthesis and Recommendations for Further Investigation

## Key Decision Points

This report has outlined analytical frameworks for three interconnected research questions. Rather than offering definitive answers, we highlight key decision points where deeper investigation is warranted:

### For RQ1 (Class-Survival Mediation)
- **Decision**: How much reduction in the class effect is "meaningful"? Is a 20% reduction sufficient to conclude that access explains the gap, or should we require >50%?
- **Investigation**: Seek additional cabin-level data (deck assignments, lifeboat capacity) to improve access proxies beyond what is available in standard passenger records.

### For RQ2 (Consistency of Rules)
- **Decision**: Are interaction effects practically significant, or do they represent noise? 
- **Investigation**: Examine historical accounts and crew testimony to validate whether different rules were indeed applied to different classes and family structures.

### For RQ3 (Missing Data Inequality)
- **Decision**: Is complete-case analysis defensible, or does it bias conclusions by excluding lower-class passengers?
- **Investigation**: Conduct all RQ1 and RQ2 analyses under multiple missing data assumptions to quantify sensitivity.

## Integration Across Questions

The three research questions are deeply linked:

- **RQ3 feeds into RQ1**: We cannot assess the role of cabin location (a key access proxy) without understanding that cabin data is missing disproportionately for 3rd class passengers.
- **RQ3 feeds into RQ2**: If age is missing for certain titles or classes, our ability to test "women and children first" for those subgroups is compromised.
- **RQ1 informs RQ2**: If class effects operate primarily through structural access, the consistency of the "women and children first" rule may vary more by deck/lifeboat proximity than by formal policy.

## Next Steps

1. **Extend the analysis** to the test set if the goal is predictive modeling, validating whether findings hold on unseen data.
2. **Deepen historical investigation** by integrating archival records, crew interviews, or other sources that document actual evacuation procedures.
3. **Refine proxies** using domain knowledge: engage historians or maritime experts to validate whether our proxy variables (cabin letter, fare, embarkation) truly reflect access constraints.
4. **Conduct robustness checks** on mediation estimates using causal inference methods (e.g., g-computation, inverse probability weighting) that relax some of the assumptions of standard logistic regression.

---

# Appendix: Technical Notes

## Software and Packages

All analyses were conducted in R using the following packages:
- `tidyverse` (1.3.x): data wrangling and visualization
- `stringr` (1.5.x): text processing (title extraction)
- `knitr` (1.45.x): table formatting

## Assumptions

1. **Mediation Analysis**: We assume that the proxies (HasCabin, UpperDeck, Fare) are correlated with true structural access. This is not verifiable from the data alone.
2. **Logistic Regression**: We assume that survival can be modeled as a Bernoulli process and that the logit link is appropriate.
3. **Missing Data**: We assume different scenarios (MAR, MCAR, MNAR) in sensitivity analyses but do not claim to know which is true.

## Reproducibility

This report is fully reproducible: all code is visible, all numbers are referenced to code outputs, and all assumptions are explicit. Future analyses can use this framework as a starting point and extend it with additional data or methods.

